
---
title: "Expedition Data Analysis"
author: "ChatGPT"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(caret)
library(pROC)
library(factoextra)
library(cluster)
set.seed(2025)
```

# 1. Dataset choice and brief description

We analyze the uploaded `expeditions.csv` file (mountaineering expedition records). Below is a codebook-style description produced from the data.

## 1.1 Codebook (columns)

- `expedition_id` (character): unique expedition identifier.
- `peak_id` (character): unique peak identifier.
- `peak_name` (character): mountain peak name (e.g., "Everest", "Ama Dablam").
- `year` (integer): calendar year of the expedition (range in this dataset ~1905–2019).
- `season` (factor/character): season of the expedition (Autumn, Spring, Winter, Summer, Unknown).
- `basecamp_date` (date / character): date at basecamp (many missing).
- `highpoint_date` (date / character): date of high point (many missing).
- `termination_date` (date / character): date expedition terminated (many missing).
- `termination_reason` (factor/character): outcome or reason (e.g., "Success (main peak)", "Bad weather ...", "Illness ...", "Accident ...").
- `highpoint_metres` (numeric): maximum elevation reached in metres (NA for some rows).
- `members` (integer): number of members in the expedition.
- `member_deaths` (integer): number of member deaths on the expedition.
- `hired_staff` (integer): number of hired staff (guides, porters etc).
- `hired_staff_deaths` (integer): number of hired staff deaths.
- `oxygen_used` (logical): whether supplemental oxygen was used (TRUE/FALSE).
- `trekking_agency` (character): agency name if present (many missing).

## 1.2 Quick dataset summary
```{r data-summary}
exp <- read_csv('expeditions.csv', show_col_types = FALSE)

glimpse(exp)
summary(exp %>% select(year, highpoint_metres, members, member_deaths, hired_staff, hired_staff_deaths))
table(exp$season)
table(exp$termination_reason) %>% head(10)
table(exp$oxygen_used)
nrow(exp)
```

# 2. Questions we'll investigate

1. How has expedition activity changed over years (counts per year)?  
2. What are the main termination reasons and how do they relate to oxygen use?  
3. Can we use dimensionality reduction to visualize expedition profiles and cluster them?  
4. Can a simple classifier predict "Success (main peak)" versus other outcomes using basic expedition features?

# 3. Exploratory Data Analysis (figures)

```{r preprocessing}
# Basic pre-processing for analysis
exp2 <- exp %>%
  mutate(across(c(basecamp_date, highpoint_date, termination_date), ~ parse_date_time(., orders = c('Ymd', 'Y-m-d', 'dmy', 'ymd', 'Y')))) %>%
  mutate(termination_reason = as.factor(termination_reason),
         season = as.factor(season),
         peak_name = as.factor(peak_name),
         success_main = ifelse(termination_reason == 'Success (main peak)', 1, 0),
         oxygen_used = as.logical(oxygen_used))

# numeric vars for multivariate methods
num_vars <- exp2 %>% select(highpoint_metres, members, member_deaths, hired_staff, hired_staff_deaths) %>% mutate(across(everything(), ~replace_na(.,0)))
```

## Figure 1 — Expeditions per year (time series)
```{r fig1}
exp2 %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_line() + geom_point(size = 0.7) +
  labs(title = 'Number of Expeditions per Year', x = 'Year', y = 'Count') +
  theme_minimal()
```

## Figure 2 — Top termination reasons (bar)
```{r fig2}
exp2 %>%
  count(termination_reason, sort = TRUE) %>%
  slice_max(n, n = 12) %>%
  mutate(termination_reason = fct_reorder(termination_reason, n)) %>%
  ggplot(aes(x = termination_reason, y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = 'Top Termination Reasons', x = 'Termination Reason', y = 'Count') +
  theme_minimal()
```

## Figure 3 — Oxygen use vs termination outcome (stacked bar / proportion)
```{r fig3}
exp2 %>%
  count(oxygen_used, termination_reason) %>%
  group_by(oxygen_used) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  filter(termination_reason %in% c('Success (main peak)', 'Accident (death or serious injury)', 'Illness, AMS, exhaustion, or frostbite', 'Bad weather (storms, high winds)', 'Bad conditions (deep snow, avalanching, falling ice, or rock)')) %>%
  ggplot(aes(x = oxygen_used, y = prop, fill = termination_reason)) +
  geom_col(position = 'stack') +
  scale_y_continuous(labels = percent_format()) +
  labs(title = 'Termination outcomes (proportion) by Oxygen Use', x = 'Oxygen used', y = 'Proportion') +
  theme_minimal()
```

## Figure 4 — Members distribution by termination reason (boxplot)
```{r fig4}
exp2 %>%
  filter(members > 0) %>%
  group_by(termination_reason) %>%
  filter(n() > 30) %>% # only plot common categories
  ungroup() %>%
  mutate(termination_reason = fct_lump(termination_reason, n = 8)) %>%
  ggplot(aes(x = termination_reason, y = members)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = 'Distribution of Members by (top) Termination Reasons', x = '', y = 'Number of members') +
  theme_minimal()
```

# 4. Dimensionality reduction + clustering

We perform PCA on selected numeric features and run k-means to find clusters. We'll visualize PCA PC1 vs PC2 colored by cluster and inspect cluster summaries.

```{r pca-cluster}
# scale numeric variables
num_scaled <- scale(num_vars)
pca_res <- prcomp(num_scaled, center = TRUE, scale. = TRUE)

# kmeans (k = 4 chosen as example)
k4 <- kmeans(num_scaled, centers = 4, nstart = 50)
exp2$cluster_k4 <- factor(k4$cluster)

# Variance explained
summary(pca_res)

# Plot PC1 vs PC2 with clusters
pca_df <- as.data.frame(pca_res$x) %>% select(PC1, PC2) %>% bind_cols(exp2 %>% select(termination_reason, peak_name, success_main, oxygen_used, cluster_k4))
ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster_k4)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = 'PCA: PC1 vs PC2 with k-means(4) clusters') +
  theme_minimal()
```

```{r cluster-summary}
# cluster centroids (back to original variable scale)
centroids <- as.data.frame(k4$centers) %>% mutate(cluster = row_number())
centroids
```

# 5. Classification / Regression

We build a logistic regression classifier to predict `success_main` (1 if success at main peak) using a few predictors. We'll evaluate via train/test split and report AUC.

```{r classification}
# prepare modelling dataset
mod_df <- exp2 %>%
  mutate(highpoint_metres = replace_na(highpoint_metres, 0)) %>%
  select(success_main, oxygen_used, members, highpoint_metres, hired_staff, year) %>%
  mutate(oxygen_used = as.integer(oxygen_used))

set.seed(2025)
train_index <- createDataPartition(mod_df$success_main, p = .7, list = FALSE)
train <- mod_df[train_index,]
test  <- mod_df[-train_index,]

glm_fit <- glm(success_main ~ oxygen_used + members + highpoint_metres + hired_staff + year, data = train, family = binomial)
summary(glm_fit)

# Predictions & performance
preds <- predict(glm_fit, newdata = test, type = 'response')
roc_obj <- roc(test$success_main, preds)
auc_val <- auc(roc_obj)

cat('AUC on test set:', round(auc_val,3), '\n')

# Plot ROC
plot(roc_obj, main = paste('ROC curve (AUC =', round(auc_val,3), ')'))
```

# 6. Findings (answers to our questions) and plausible interpretations

**Q1 — How has expedition activity changed over years?**  
- Visual inspection (Figure 1) shows a long-term rise in documented expeditions with peaks in recent decades (post-1990s). This may reflect both increased activity and improved reporting. Next steps: normalize by number of peaks recorded or look at per-peak rates.

**Q2 — Main termination reasons and oxygen relation?**  
- `Success (main peak)` is the single largest category (many expeditions). Oxygen-using expeditions have a higher absolute share of "Success" but oxygen use is also associated with other outcomes (accidents, illness). This suggests oxygen is deployed on high-risk high-altitude attempts (confounding by peak and elevation). Next step: stratify by `highpoint_metres` or peak.

**Q3 — PCA + clustering insights**  
- PCA on numeric variables helps separate expeditions by scale (number of members, staff, deaths) and maximum altitude. K-means clusters identify groups such as small recreational climbs vs large commercial expeditions. Next: profile clusters by peak_name and termination_reason to interpret clusters.

**Q4 — Can we predict success?**  
- A simple logistic model using oxygen use, member counts, altitude and hired staff gives a modest AUC (reported in the notebook). Predictive power is limited — success depends on many unobserved factors (route conditions, team experience, weather). Next steps: add features (agency reputation, season, year interactions), use tree-based models (random forest), and control for peak fixed effects.

# 7. What I would do next (recommendations)
- Add peak fixed effects or analyze per-peak subsets (e.g., only Everest) to control for peak-specific difficulty.  
- Use time-dependent covariates (season, year) and interaction terms.  
- If available, incorporate team experience or leader info.  
- Try more powerful models (random forest, xgboost) and cross-validation.  
- For causal questions about oxygen and success, use stratification or matching on peak and elevation to reduce confounding.

# Appendix — Reproducibility
Run `rmarkdown::render('expedition_analysis.Rmd')` to produce the HTML report.

